---
title: "Describing the Cleaned NHANES Dataset"
author: "Charles-Antoine Auger"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  pdf_document:
    number_sections: true
    latex_engine: pdflatex
bibliography: references.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

library(pacman)

pacman::p_load(
  here,       # project-relative file paths
  rio,        # data import
  tidyverse,  # data manipulation
  knitr,      # kable()
  kableExtra, # table formatting
  scales      # number formatting
)

nhanes <- import(here("data", "cleaned_NHANES.csv"))

sbp_cols <- c("systolic_bp_1", "systolic_bp_2",
              "systolic_bp_3", "systolic_bp_4")

nhanes <- nhanes |>
  mutate(
    wave    = factor(wave_id),
    sbp_avg = rowMeans(across(all_of(sbp_cols)), na.rm = TRUE),
    sbp_avg = if_else(is.nan(sbp_avg), NA_real_, sbp_avg)
  )
```

# Introduction

The National Health and Nutrition Examination Survey (NHANES) is a continuous
programme of the National Center for Health Statistics that combines
interviews, physical examinations and laboratory measurements to describe the
health and nutritional status of the civilian, non-institutionalised population
of the United States [@nhanes]. Participants are selected through a complex
multistage probability design, and data are released in two-year waves.

This report describes the cleaned NHANES extract used throughout BIOS 640. The
extract contains `r comma(nrow(nhanes))` participants across three waves and
retains demographic variables, four repeated systolic and diastolic blood
pressure readings, and a small set of socioeconomic indicators. Two tables
summarise the dataset: the first covers sample composition, the second the
distribution of blood pressure. All analyses were carried out in R [@rcore],
and the tables were formatted with kableExtra [@kableextra].

# Sample composition

```{r table-composition}
n_by_wave <- nhanes |>
  count(wave, name = "n_total")

size_row <- n_by_wave |>
  mutate(level = "Participants", cell = comma(n_total)) |>
  select(wave, level, cell)

sex_row <- nhanes |>
  group_by(wave) |>
  summarise(n = sum(gender == "Male"), .groups = "drop") |>
  left_join(n_by_wave, by = "wave") |>
  mutate(level = "Male",
         cell  = sprintf("%s (%.1f)", comma(n), 100 * n / n_total)) |>
  select(wave, level, cell)

eth_rows <- nhanes |>
  count(wave, ethnicity_2, name = "n") |>
  left_join(n_by_wave, by = "wave") |>
  mutate(cell = sprintf("%s (%.1f)", comma(n), 100 * n / n_total)) |>
  select(wave, level = ethnicity_2, cell)

lev_order <- c("Participants", "Male", "White", "Black", "Mexican",
               "Asian", "Other Hispanic", "Other")

tab1 <- bind_rows(size_row, sex_row, eth_rows) |>
  mutate(level = factor(level, levels = lev_order)) |>
  arrange(level) |>
  pivot_wider(names_from = wave, values_from = cell) |>
  rename(` ` = level)

kbl(tab1,
    booktabs = TRUE,
    align    = c("l", "c", "c", "c"),
    caption  = "Composition of the cleaned NHANES sample by survey wave") |>
  kable_styling(latex_options = c("hold_position"), font_size = 10) |>
  add_header_above(c(" " = 1, "NHANES wave" = 3)) |>
  pack_rows("Sample size", 1, 1) |>
  pack_rows("Sex, n (%)", 2, 2) |>
  pack_rows("Race and Hispanic origin, n (%)", 3, 8) |>
  column_spec(1, width = "4.2cm") |>
  row_spec(1, bold = TRUE) |>
  footnote(
    general = paste("Percentages are column percentages, computed within each",
                    "wave. Sex reflects the original NHANES variable, which",
                    "records biological sex rather than gender identity."),
    general_title  = "Note:",
    threeparttable = TRUE
  )
```

The three waves are of comparable size, ranging from 9,254 to 10,175
participants. The sex distribution is remarkably stable: males represent
between 49.1% and 49.2% of each wave, so any difference observed across waves
cannot be attributed to a shift in sex composition.

Race and Hispanic origin vary more. The White group is the largest in every
wave, but its share moves by more than five percentage points, from 36.1% in
2015 to 30.7% in 2016. Other Hispanic participants show the opposite pattern,
peaking at 13.1% in 2016 before falling back to 8.9% in 2017. These movements
appear to reflect the oversampling strategy NHANES applies to selected
subgroups, which changes between survey cycles, rather than a genuine change in
the underlying population.

# Blood pressure

```{r table-bp}
tab2 <- nhanes |>
  filter(!is.na(sbp_avg)) |>
  group_by(wave) |>
  summarise(
    n      = n(),
    Min    = min(sbp_avg),
    Median = median(sbp_avg),
    Mean   = mean(sbp_avg),
    SD     = sd(sbp_avg),
    Max    = max(sbp_avg),
    .groups = "drop"
  ) |>
  mutate(across(Min:Max, \(x) sprintf("%.1f", x)),
         n = comma(n)) |>
  rename(Wave = wave)

kbl(tab2,
    booktabs = TRUE,
    align    = c("l", rep("c", 6)),
    caption  = paste("Average systolic blood pressure by survey wave, among",
                     "participants with at least one valid reading")) |>
  kable_styling(latex_options = c("hold_position"), font_size = 10) |>
  add_header_above(c(" " = 2,
                     "Average systolic blood pressure (mm Hg)" = 5)) |>
  column_spec(5, bold = TRUE) |>
  footnote(
    general = paste("The average systolic blood pressure is the mean of up to",
                    "four readings taken at a single examination, and is",
                    "missing only when all four readings are unavailable.",
                    "Roughly one participant in four has no valid reading",
                    "(26.0%, 26.2% and 27.5% across the three waves). The mean",
                    "is shown in bold to highlight the upward trend."),
    general_title  = "Note:",
    threeparttable = TRUE
  )
```

Average systolic blood pressure rises steadily across the three waves, from
118.3 mm Hg in 2015 to 120.4 in 2016 and 121.7 in 2017, a total increase of
3.4 mm Hg. The medians move in the same direction, which suggests that the
shift affects the bulk of the distribution rather than being driven by extreme
values alone.

The spread also widens, with the standard deviation increasing from 18.1 to
20.3 mm Hg. The maximum values sit close to 230 mm Hg in every wave, so the
upper tail is not what changed; rather, more participants fall further from the
centre in the later wave. Because the waves differ in their race and Hispanic
origin composition, and because age is a strong determinant of blood pressure,
this trend should be read as a description of these samples rather than as
evidence of a change in the underlying population.

# References
